Progress Report for CSS 586 Course Project: \\ Modeling Latent Patterns in Music

### 2021-05-01 papers/progress/alex/progress-alex.tex
save work on MIDI pianoroll processing functions

MusicVAE, part of the Magenta project, is a variational LSTM
autoencoder for MIDI that incorporates a novel hierarchical structure
using a ``conductor'' recurrent layer in its decoder model to better
capture structure at multiple levels and avoid the problem of
``posterior/mode collapse'' whereby a generative model learns to
ignore its latent code and rely on autoregression
\cite{roberts_hierarchical_2018}.

@@

Other research has focused on modeling raw audio waveforms directly. WaveNet is
a causal convolutional neural network for generating raw audio waveforms,
developed by Google DeepMind, which achieves state of the art performance in
generating natural sounding speech from text, but is also capable of generating
short, realistic snippets of audio music \cite{oord_wavenet_2016}.

Another model named SampleRNN generates raw audio waveforms using a three-tier
hierarchy of gated recurrent units (GRU) to model recurrent structure at
multiple temporal resolutions \cite{mehri_samplernn_2017}.

@@

Two primary datasets will be used for this research project: MusicNet,
which is a collection of 330 freely licensed European classical music
recordings with aligned MIDI scores \cite{thickstun2017learning}, and
the Lakh MIDI Dataset, which includes a collection of 45,129 Creative
Commons licensed MIDI files of popular music songs that have been
matched and aligned to MP3 recordings, and of which we will use a
subset for model training and evaluation
\cite{raffel_learning-based_2016}.

@@

Several choices must be made in how to preprocess binary MIDI files
into training examples for a neural network. There are multiple
open-source Python packages capable of reading MIDI files into NumPy
array-based representations, such as pretty\_midi
\cite{raffel_pretty_midi_2014} and Pypianoroll
\cite{dong_pypianoroll_2018}. In order to accommodate polyphonic
music, we will convert each MIDI file into a pianoroll representation,
wherein each instrument track is a sparse matrix that multi-hot
encodes the velocity values for each of 128 possible pitch levels at
each timestep. Because songs are typically at least a few minutes long
and of varying length, it will not be feasible to train with entire
songs as examples, so we will crop songs into phrases of equal numbers
of measures, to use as training data.

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{first_four_bars.png}
  \caption{Pianoroll visualization of the first four measures of
    Beethoven's Serioso String Quartet}
  \Description{A pianoroll representation of cello, viola and violin
    parts from a Beethoven piece.}
\end{figure}

>>>>

MusicVAE, part of the Magenta project, is a variational LSTM
autoencoder for MIDI that incorporates a novel hierarchical structure
using a ``conductor'' recurrent layer in its decoder model to better
capture structure at multiple levels and avoid the problem of
``posterior/mode collapse'' whereby a generative model learns to
ignore its latent code and rely on autoregression
\cite{roberts_hierarchical_2018}. This model is capable of generating
new melodies that blend two given melodies via latent space
interpolation.

@@

Other research has focused on modeling raw audio waveforms directly. WaveNet is
a causal convolutional neural network for generating raw audio waveforms,
developed by Google DeepMind, which achieves state of the art performance in
generating natural sounding speech from text, but is also capable of generating
short, realistic snippets of audio music \cite{oord_wavenet_2016}.
Another model named SampleRNN generates raw audio waveforms using a three-tier
hierarchy of gated recurrent units (GRU) to model recurrent structure at
multiple temporal resolutions \cite{mehri_samplernn_2017}.

@@

Two primary datasets will be used for this research project: MusicNet,
which is a collection of 330 freely licensed European classical music
recordings with aligned MIDI scores \cite{thickstun2017learning}, and
the Lakh MIDI Dataset, which includes a collection of 45,129 Creative
Commons licensed MIDI files of popular music songs that have been
matched and aligned to MP3 recordings, and of which we will use a
subset for model training and evaluation
\cite{raffel_learning-based_2016}. Including this second dataset may
help us to generalize the modeling approach beyond European classical
music to include other popular genres and associated instruments.

@@

Several choices must be made in how to preprocess binary MIDI files
into training examples for a neural network. There are multiple
open-source Python packages capable of reading MIDI files into NumPy
array-based representations, such as pretty\_midi
\cite{raffel_pretty_midi_2014} and Pypianoroll
\cite{dong_pypianoroll_2018}. In order to accommodate polyphonic
music, we will convert each MIDI file into a pianoroll representation,
as visualized in Figure \ref{pianoroll}, wherein each instrument track
is a sparse matrix that multi-hot encodes the velocity values for each
of 128 possible pitch levels at each timestep. Because songs are
typically at least a few minutes long and of varying length, it will
not be feasible to train with entire songs as examples, so we will
crop songs into phrases of equal numbers of measures to use as
training data.

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{first_four_bars.png}
  \caption{Pianoroll visualization of the first four measures of
    Beethoven's Serioso String Quartet}
  \label{pianoroll}
  \Description{A pianoroll representation of cello, viola and violin
    parts from a Beethoven piece.}
\end{figure}
